The Right Way to Search Evolving Graphs

### 2016-02-22 paper.tex
One last round of changes

\usepackage{url}
% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

@@

\begin{abstract}
Evolving graphs arise in many different contexts where the
interrelations between data elements change over time.
We present a breadth first search (BFS) algorithm for evolving graphs
that can track (active) nodes correctly.
Using simple examples, we show that na\"ive matrix-matrix multiplication on
time-dependent adjacency matrices miscounts the number of temporal paths.
By mapping an evolving graph to an adjacency matrix of the equivalent static graph,
we prove the properties of the BFS algorithm using the properties of the
adjacency matrix. Finally, demonstrate how the BFS over evolving graphs can be
applied to mining citation networks.

@@

Although it is well known that sparse matrix-vector product is equivalent to a 
one-step BFS on the corresponding graph~\cite{kegi11},
it is little known about the generalization of this duality for evolving graphs.
In fact, previous works form the off-diagonal blocks wrongly~\cite{fehi15}. 
We prove the generalization of this duality by finding
the correct block sparse matrix representation of an evolving graph.
This work forms the basis for many algebraic evolving graph algorithms.

@@

This definition of a temporal path differs from that of the dynamic walk
in \cite{gphe11,grihig13} in its explicit requirement of containing only active
nodes. The definition implies that if either or both end points of a temporal
path are inactive, then the entire temporal path must be the empty sequence
$\langle \rangle$. Keeping track explicitly of the time labels of each temporal
node allows greater generality to cases where the node sets change over time.
Furthermore, we shall show later in Sec.~\ref{sec:temp-paths-adjac}
that the explicit bookkeeping of the time labels is essential for correctly
generalizing the BFS to evolving graphs.

@@

@@

The BFS on evolving graphs is described in Algorithm~\ref{alg:bfs}.
Algorithm~\ref{alg:bfs} is identical to the BFS on static graphs expect 
line $8$, where we visit the forward neighbors of a given temporal node
in both space and time.
Given an evolving graph $G_n$ and a root $(v_1, t_1)$,
Algorithm~\ref{alg:bfs} returns all temporal notes reachable from the root and
their distances from the root.
$reached$ is a dictionary from temporal nodes to integers whose key set represents
all visited temporal nodes and whose value set are the corresponding distances
from the root.

@@

Similarly, define an edge set $E' = \{(u_s,v_t)|u_s=(u,s)\in V,
v_t=(v,t)\in V, v=u, s<t\}$, which consists of temporal nodes that connect
active nodes sharing the same node at different times. Each edge in $E'$ is then
in 1-1 correspondence with a temporal path of length 2,
$\langle (v,s), (v,t) \rangle$.
Define also $\tilde E^{[t]} = \{(e, t)|e\in E^{[t]}\}$, which are simply the edge
sets in $G_n$ with time labels. Then $E = \bigcup_t \tilde E^{[t]} \cup E'$ is
the set of all edges representing all allowed temporal paths of length 2.

@@

We now establish a similar 1-1 correspondence of forward neighbors of an active
node with a subset of $G$. By induction, we can show that at iteration $k$,
all new nodes populated into the key set of $reached$ by definition are of
distance $k$ from the root. By definition,
the forward neighbors of some active node $(v, t) \in G_n$ are
active nodes of either the form $(v, t')$ for some $t' > t$ or $(u, t)$ for some
$u\ne v$. Clearly, the former are elements of $E' \subseteq E$ while the latter
are elements of $E^{[t]} \subseteq \bigcup_s \tilde E^{[s]} \subseteq E$.
Thus each forward neighbor of an active node $(v, t) \in G_n$ is in 1-1
correspondence with a node in $V$ that is a neighbor of $v_t \in V$.

@@

\begin{thm}[Complexity of the evolving graph BFS]
\label{thm:complexity}
Let $G_n$ be an evolving graph represented using adjacency lists,
$(v_1, t_1)$ be an active node of $G_n$, and $G=(V,E)$ be the static graph
constructed from $G_n$ using the 1-1 correspondences defined in the proof of
Theorem~\ref{thm:correct}. Then the asymptotic complexity of
Algorithm~\ref{alg:bfs} is $O(|E| + |V|)$.
\end{thm}

@@

Note the edge set $E$ is $\cup_t\tilde E^{[t]} \cup E'$, which contains edges that
are causal structured.

@@

\subsection{Not all temporal paths are counted by products of adjacency matrices}
\label{sec:temp-paths-adjac}

For each static graph $G^{[t]} = (V^{[t]}, E^{[t]})$ that constitutes the
evolving graph $G_n$, define its corresponding
$\left\vert V^{[t]}\right\vert \times \left\vert V^{[t]}\right\vert$
adjacency matrix with elements
\begin{equation}
A^{[t]}_{ij} =
\begin{cases}
 1 & \mbox{if $(i,j)\in E^{[t]}$,} \\
 0 & \mbox{otherwise.}
\end{cases}
\label{eqn:adjacency}
\end{equation}

@@

Sums $S^{[t]}$ of the form \eqref{eq:sum} produce an incorrect count of temporal
paths because they do not capture temporal paths with subpaths of the form
$\langle (v, s), (v, t) \rangle$, $s < t$.
One might attempt to amend the sums $S^{[t]}$ in \eqref{eq:sum} by redefining
the adjacency matrices to include ones along the diagonal, hence allowing
paths containing the sequence $\langle (i, t_1), (i, t_2) \rangle$. However, the resulting
sum is still incorrect, as it counts paths with subsequences $\langle (3, t_1), (3, t_2) \rangle$
and are hence not temporal paths. Instead, the temporal path \eqref{eq:tpathex}
is counted by the matrix product $M^{[t_1, t_2]} A^{[t_2]}$, where
\begin{equation}
    M^{[t_1, t_2]} = \begin{pmatrix}1 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & 0\end{pmatrix}.
\label{eq:m-example}
\end{equation}
$M^{[t_1, t_2]}$ describes the forward time propagation of temporal nodes that
are active at both times $t_1$ and $t_2$, i.e.\ it counts
temporal paths that contain subsequences $\langle (i, t_1), (i, t_2) \rangle$,
and both $(i, t_1)$ and $(i, t_2)$ are active nodes.

The simple example of Figure~\ref{fig:eg_shortest_path} provides several
counterexamples which demonstrate why sums over products of adjacency matrices of
the form \eqref{eq:sum} do not count temporal paths correctly. The interpretation
from considering powers of $A$ for static graphs therefore does not extend to
evolving graphs, and sums like \eqref{eq:sum} cannot form the basis of correct
analyses of evolving graphs.

@@

\[
A^T \!\odot b =
\begin{cases}
b & \mbox{if $A^Tb \ne 0$ or $Ab \ne 0$,} \\
0 & \mbox{otherwise.}
\end{cases}
\]
The forward neighbors of a temporal node $(k, t_1)$ in $A_n$ can then be determined
from the indices and time stamps of the nonzero elements in the sequence
\begin{equation}
\label{eq:out_ne}
\big\langle (A^{[1]})^Te_k, \; (A^{[2]})^T\!\odot e_k, \; \ldots \; (A^{[n]})^T\!\odot e_k\big\rangle.
\end{equation}
The nonzero entries of the first vector represent forward neighbors that are on
the same time stamp $t_1$, whereas nonzero entries of the other vectors represent forward
neighbors that are advanced in time but remain on the same node $k$.
The quantity \eqref{eq:out_ne} therefore encodes a BFS tree of depth 2,
as its nonzero entries are labeled by all temporal nodes of distance 1 from
$(k, t_1)$.

@@

\[
\bm M_n = \begin{bmatrix}
A^{[t_1]} & M^{[t_1, t_2]} & ... & M^{[t_1, t_n]} \\
        0 & A^{[t_2]}      & ... & M^{[t_2, t_n]} \\
                           &   ... \\
        0 & 0 & ... & A^{[t_n]}
\end{bmatrix}
\]
%
where $M^{[t_i, t_j]}$ is the matrix whose rows are labeled by $V^{[t_i]}$ and
columns are labeled by $V^{[t_j]}$, and whose entries are
\[
M^{[t_i, t_j]}_{uv} =
\begin{cases}
1 & \mbox{if $(u,v)\in E^\prime$,} \\
0 & \mbox{otherwise.}
\end{cases}
\]
%
The adjacency matrix blocks $A^{[t]}$ encode the edge set $\tilde E^{[t]}$,
whereas the off-diagonal blocks $M^{[t_i, t_j]}$ together encode the edge set
$E^\prime$ which capture temporal paths with subsequences of the form
$\langle (v, t_i), (v, t_j) \rangle$. Then $\bm M_n$ is the adjacency matrix of
the graph $(\cup_t V^{[t]}, E)$, which is the graph $G$ together with all the
inactive nodes. From the definition, $\bm M_n$ has nonzero
entries only in rows and columns that correspond to active nodes $V$, and so
retaining only these rows and columns corresponding to $V$ produces the
adjacency matrix $\bm A_n$ of $G = (V, E)$.

@@

Finally, we prove a simple lemma that the block adjacency matrix $\bm A_n$ is
nilpotent whenever all the subgraphs $G^{[t]}$ of $G_n$ are acyclic.

@@

\IEEEproof
Since each directed graph $G^{[t]}$ is acyclic, its corresponding adjacency
matrix $A^{[t]}$ is strictly upper triangular. As a result, $\bm A_n$ must be
upper triangular.

@@

The blocked matrix structure of the adjacency matrices presented here provide
interesting relationships between their matrix properties and the algorithmic
properties of BFS, made possible because of the reformulation of BFS as
repeated power iterations of the adjacency matrix in Algorithm~\ref{alg:bfsla}.
Note, however, that these matrices never need to be instantiated for practical
computations. Rather, since Algorithm~\ref{alg:bfsla} only requires the
matrix--vector product involving the adjacency matrix, the formulation of
Algorithm~\ref{alg:bfs} provides an efficient way to exploit the block structure
of $\bm A_n$. The $\odot$ operation provides an efficient way to compute the
action of the off-diagonal products. Representing the diagonal blocks
$A^{[t]}$ as sparse matrices product further reduces the cost of BFS by exploiting
latent sparsity in graphs that show up in practical applications.

@@

\IEEEproof
Recall from Lemma~\ref{thm:nilpotent} that $\bm A_n$ is nilpotent, i.e.\
there exists some positive integer $k$ for which $\bm A_n^k = 0$.
Hence, after iteration $k$, $\bm b$ is assigned the value $\bm A_n^k \bm b = 0$.
Therefore, Algorithm~\ref{alg:bfsla} must terminate after iteration $k$.
\endproof

@@

Given an evolving graph $G_n = \langle G^{[1]}, G^{[2]}, \ldots, G^{[n]} \rangle$, let $V$ be the set of active nodes,
$E = \cup_t \tilde E^{[t]} \cup E'$ be the set of edges, where
$\tilde E^{[t]}$ represents the set of edges of $G^{[t]}$ and $E'$ represents the set
of causal edges between time stamps. Let $\bm A_n$ be the corresponding
adjacency matrix of $G_n$.

Let's analyze the running time of Algorithm \ref{alg:bfsla}. 
Suppose $\bm A_n$ is dense.
Forming the block adjacency matrix $\bm A_n^T$ in line $3$
requires $\Theta(|V|^2)$ operations. 
We note, however, $\bm A_n^T$ need never be formed
explicitly, as only its matrix-vector product is required in the algorithm.
The matrix-vector product in line $7$ requires $O(|V|^2)$ operations.
In the worst case, we need to reach all active nodes and reach only one 
active node at each iteration. 
So the number of iterations $k$ of the while loop at line $6$ is bounded by
$|E|$, i.e., $k = \Theta(|E|)$. Hence, the total running time is 
$\Theta(|E||V|^2)$. This is much worse than Algorithm \ref{alg:bfs}, which
has linear running time.

If $\bm A_n$ is sparse, forming $\bm A_n$ costs $\Theta(|E|)$ flops. 
The gaxpy operation in line $7$ costs 
$2m$ flops, where $m$ is the number of none zeros of $\bm A_n$ \cite[Sec.11]{golo13}. 
Here $m$ is equal to $|E|$. Thus, in the worst case, the total running time is 
$\Theta(|E|^2)$.

@@

To study evolving graphs and experiment with various graph types, we have
developed EvolvingGraphs.jl~\cite{zhang15}, a software package for the creation,
manipulation, and study of evolving graphs written in Julia~\cite{bkse12}.
It is freely available online at
\begin{center}
\url{https://github.com/weijianzhang/EvolvingGraphs.jl}
\end{center}
and available with the MIT ``Expat'' license. The package contains an implementation
of the evolving graph BFS. \texttt{IntEvolvingGraph}, a data type in EvolvingGraphs.jl,
represents an evolving graph as adjacency lists.
We generate a group of  random (directed) \texttt{IntEvolvingGraph}s
with $8,000$ active nodes and $4$ time stamps.
At each time stamp $i$, an edge is included in $G^{[i]} \in G_4$ with probability $p$.
 Figure \ref{fig:time} shows the plots of  number of edges against the computation time
for running Algorithm \ref{alg:bfs}
in Julia. All experiments are conducted 
in a Linux system with Intel(R) Core(TM) i7-3537U CPU @ $2.00$GHz.
The results show Algorithm \ref{alg:bfs} can be computed in linear time, which agrees
with Theorem \ref{thm:complexity}. A summary of the experiments is provided in Table \ref{tab:bfs}.
\begin{figure}[h]
  \centering
  \includegraphics[scale=0.44]{time.pdf}
  \caption{Experiments of running Algorithm \ref{alg:bfs}
on a group of random evolving graphs with $8,000$ active nodes and $4$ time stamps. The x-axis shows the
number of edges of each evolving graph and the y-axis shows the corresponding
computation time. The values in the x, y-axis increase linearly from left to 
right and from bottom to top respectively.}
  \label{fig:time}
\end{figure}
\begin{table}[h]
\centering
\begin{tabular}{c | c | c | c }
\hline
probability $p$ & no. of edges  & time (s) &  memory (MB) \\\hline
      0.01  &         160,156      & 0.098   & 18.388  \\
      0.09  &       1,440,895        & 0.446  &  134.253  \\
      0.15   &     2,399,774         & 0.743 & 221.511 \\
      0.23   &    3,676,959          & 1.206 &   337.688 \\
      0.31   &     4,956,969          & 1.504  &  494.833 \\
\end{tabular}
\caption{the computation time and the amount of memory allocated for 
computing Algorithm \ref{alg:bfs}. This is a summery of Figure \ref{fig:time}.  
Each edge is included in the graph with probability $p$. }
\label{tab:bfs}
\end{table}
%Julia is a high-level, high-performance dynamic programming language for technical
%computing.
% We make particular use of
% the parameterizable type system and multiple dispatch in Julia.
% \texttt{IntMatrixList}, a data type in EvolvingGraphs.jl, represents an evolving graph
% as a sequence of sparse matrices stored in compressed sparse column (CSC) format.
% We can convert an \texttt{IntMatrixList} to its corresponding block adjacency matrix
% using the function \texttt{spmatrix}.

@@

In this section, we show that evolving graph formalism presented above can be
used to capture the structure of citation networks. Consider
the evolving graph $G_n=\langle G^{[t]} \rangle_t$ such that $G^{[t]}$ has
node set corresponding to authors active at time $t$ and directed edge set
$E^{[t]} \ni (i, j)$ representing a citation of author $j$ by author $i$ in a
publication at time $t$.

@@

>>>>

\usepackage{url}
% correct bad hyphenation here
\hyphenation{}

@@

\begin{abstract}
Evolving graphs arise in problems where interrelations between data change over
time. We present a breadth first search (BFS) algorithm for evolving graphs
that computes the most direct influences between nodes at two different times.
Using simple examples, we show that na\"ive unfoldings of adjacency matrices
miscount the number of temporal paths.
By mapping an evolving graph to an adjacency matrix of an equivalent static graph,
we prove that our generalization of the BFS algorithm is correct.
Finally, we demonstrate how the BFS over evolving graphs can be applied to mine
citation networks.

@@

It is well known that sparse matrix-vector product is equivalent to a
one-step BFS on the corresponding graph~\cite[Sec. 1.1]{kegi11}.
However, little is known about the generalization of this duality for evolving graphs.
We prove the generalization of this duality by finding
the correct block sparse matrix representation of an evolving graph.

@@

This definition of a temporal path differs from that of the dynamic walk
in \cite{gphe11,grihig13} in that \textbf{causal edges}, i.e.\ sequences of the
form $\langle (v, t), (v, t\prime)\rangle$ are included explicitly in temporal
paths but are only implicitly included in dynamic walks and are not counted
toward the length of dynamic walks.
Our definition implies that if either or both end points of a temporal
path are inactive, then the entire temporal path must be the empty sequence
$\langle \rangle$. Keeping track explicitly of the time labels of each temporal
node allows greater generality to cases where the node sets change over time.
Furthermore, we shall show later in Sec.~\ref{sec:temp-paths-adjac}
that the explicit bookkeeping of the time labels is essential for correctly
generalizing the BFS to evolving graphs.

@@

Our definition of distance, again, differs from the definition of distance
in the formulation of \cite{gphe11,grihig13} in that we explicitly count causal
edges toward the distance. It also differs from the notion of temporal distance
in the work of Tang and coworkers~\cite{tmml10}, which is the number of time
steps between $t$ and $s$ (inclusive). In this respect, our formulation of the
BFS on evolving graphs differs from these other works by minimizing a different
notion of distance over an evolving graph.

@@

The BFS on evolving graphs is described in Algorithm~\ref{alg:bfs}.
Algorithm~\ref{alg:bfs} is identical to the BFS on static graphs except for
line $8$, where we visit the forward neighbors of a given temporal node
in both space and time.
Given an evolving graph $G_n$ and a root $(v_1, t_1)$,
Algorithm~\ref{alg:bfs} returns all temporal notes reachable from the root and
their distances from the root.
$reached$ is a dictionary from temporal nodes to integers whose key set represents
all visited temporal nodes and whose value set are the corresponding distances
from the root.

@@

Similarly, define the set of \textbf{causal edges} $E' = \{(u_s,v_t)|u_s=(u,s)\in V,
v_t=(v,t)\in V, v=u, s<t\}$, which consists of temporal nodes that connect
active nodes sharing the same node at different times. Each edge in $E'$ is then
in 1-1 correspondence with a temporal path of length 2,
$\langle (v,s), (v,t) \rangle$.
Define also the set of \textbf{static edges at time $t$},
$\tilde E^{[t]} = \{(e, t)|e\in E^{[t]}\}$, which are simply the edge
sets in $G_n$ with time labels, and the set of \textbf{static edges} $\tilde E$,
being simply the union over all times, $\bigcup_t \tilde E^{[t]}$.
Then $E = \tilde E \cup E'$ is the set of all edges representing all allowed
temporal paths of length 2.

@@

We now establish a similar 1-1 correspondence of forward neighbors of an active
node with a subset of $G$. By induction, all new nodes populated into the key
set of $reached$ at iteration $k$ are of distance $k$ from the root. By definition,
the forward neighbors of some active node $(v, t) \in G_n$ are
active nodes of either the form $(v, t')$ for some $t' > t$ or $(u, t)$ for some
$u\ne v$. In other words, they are connected either by a causal edge or a static edge.
Clearly, the former are elements of $E' \subseteq E$ while the latter
are elements of $\tilde E \subseteq E$.
Thus each forward neighbor of an active node $(v, t) \in G_n$ is in 1-1
correspondence with a node in $V$ that is a neighbor of $v_t \in V$.

@@

\begin{thm}[Computational complexity of the evolving graph BFS]
\label{thm:complexity}
Let $G_n$ be an evolving graph represented using adjacency lists,
$(v_1, t_1)$ be an active node of $G_n$, and $G=(V,E)$ be the static graph
constructed from $G_n$ using the 1-1 correspondences defined in the proof of
Theorem~\ref{thm:correct}. Then the asymptotic computational complexity of
Algorithm~\ref{alg:bfs} is $O(|E| + |V|)$.
\end{thm}

@@

Note that in the theorems in this section construct an equivalent static graph
$G$ corresponding to the evolving graph $G_n$. However, $G$ contains more edges
than the union of all the static parts of $G_n$, as we also add causal edges
$E'$. To our knowledge, our formulation of the BFS represents the first attempt
to include these edges explicitly in the treatment of evolving graphs.

@@

\subsection{The importance of causal edges}
\label{sec:temp-paths-adjac}

For each static graph $G^{[t]} = (V^{[t]}, E^{[t]})$ that constitutes the
evolving graph $G_n$, define its corresponding
$\left\vert V^{[t]}\right\vert \times \left\vert V^{[t]}\right\vert$
one-sided adjacency matrix with elements
\begin{equation}
A^{[t]}_{ij} =
\begin{cases}
 1 & \mbox{if $(i,j)\in E^{[t]}$,} \\
 0 & \mbox{otherwise.}
\end{cases}
\label{eqn:adjacency}
\end{equation}

@@

Sums $S^{[t]}$ of the form \eqref{eq:sum} produce an incorrect count of temporal
paths because they do not capture temporal paths with causal edges, i.e.\ subpaths
of the form $\langle (v, s), (v, t) \rangle$, $s < t$.
One might attempt to amend the sums $S^{[t]}$ in \eqref{eq:sum} by redefining
the adjacency matrices to include ones along the diagonal, hence allowing
paths containing the sequence $\langle (i, t_1), (i, t_2) \rangle$. However, the resulting
sum is still incorrect, as it counts paths with subsequences $\langle (3, t_1), (3, t_2) \rangle$
and are hence not temporal paths. Instead, the temporal path \eqref{eq:tpathex}
is counted by the matrix product $M^{[t_1, t_2]} A^{[t_2]}$, where
\begin{equation}
    M^{[t_1, t_2]} = \begin{pmatrix}1 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & 0\end{pmatrix}.
\label{eq:m-example}
\end{equation}
$M^{[t_1, t_2]}$ describes the forward time propagation of temporal nodes that
are active at both times $t_1$ and $t_2$, i.e.\ it counts
temporal paths that contain subsequences $\langle (i, t_1), (i, t_2) \rangle$,
and both $(i, t_1)$ and $(i, t_2)$ are active nodes.

The simple example of Figure~\ref{fig:eg_shortest_path} provides several
counterexamples which demonstrate why sums over products of adjacency matrices of
the form \eqref{eq:sum} do not count temporal paths correctly. The interpretation
from considering powers of $A$ for static graphs therefore does not extend to
evolving graphs, and sums like \eqref{eq:sum} cannot form the basis of correct
analyses of evolving graphs as they neglect the combinatorics associated with
the causal edge set $E'$.

@@

\[
A^T \!\odot b =
\begin{cases}
b & \mbox{if $A^Tb \ne 0$ or $Ab \ne 0$,} \\
0 & \mbox{otherwise.}
\end{cases}
\]
The condition $A^Tb \ne 0$ ensures that the product is nonzero in components
involving left active nodes $\cup_t \tilde V_L^{[t]}$, and the condition
$Ab \ne 0$ is the analogue for right active nodes $\cup_t \tilde V_R^{[t]}$.
The forward neighbors of a temporal node $(k, t_1)$ in $A_n$ can then be determined
from the indices and time stamps of the nonzero elements in the sequence
\begin{equation}
\label{eq:out_ne}
\big\langle (A^{[1]})^Te_k, \; (A^{[2]})^T\!\odot e_k, \; \ldots \; (A^{[n]})^T\!\odot e_k\big\rangle.
\end{equation}
The nonzero entries of the first vector represent forward neighbors that are on
the same time stamp $t_1$, whereas nonzero entries of the other vectors represent forward
neighbors that are advanced in time but remain on the same node $k$.
The quantity \eqref{eq:out_ne} therefore encodes a BFS tree of depth 2,
as its nonzero entries are labeled by all temporal nodes of distance 1 from
$(k, t_1)$.

@@

\[
\bm M_n = \begin{bmatrix}
A^{[t_1]} & M^{[t_1, t_2]} & ... & M^{[t_1, t_n]} \\
        0 & A^{[t_2]}      & ... & M^{[t_2, t_n]} \\
                           &   ... \\
        0 & 0 & ... & A^{[t_n]}
\end{bmatrix}
\]
%
where $M^{[t_i, t_j]}$ is the matrix whose rows are labeled by $V^{[t_i]}$ and
columns are labeled by $V^{[t_j]}$, and whose entries are
\[
M^{[t_i, t_j]}_{uv} =
\begin{cases}
1 & \mbox{if $(u,v)\in E^\prime$,} \\
0 & \mbox{otherwise.}
\end{cases}
\]
%
The adjacency matrix blocks $A^{[t]}$ encode the static edge set $\tilde E$,
whereas the off-diagonal blocks $M^{[t_i, t_j]}$ together encode the causal edge
set $E^\prime$, which capture temporal paths with subsequences of the form
$\langle (v, t_i), (v, t_j) \rangle$. Then $\bm M_n$ is the adjacency matrix of
the graph $(\cup_t V^{[t]}, E)$, which is the graph $G$ together with all the
inactive nodes. From the definition, $\bm M_n$ has nonzero
entries only in rows and columns that correspond to active nodes $V$, and so
retaining only these rows and columns corresponding to $V$ produces the
adjacency matrix $\bm A_n$ of $G = (V, E)$.

@@

Finally, we note that some results regarding the BFS on evolving graphs can be
derived easily using properties of the block adjacency matrix $\bm A_n$. For
example, we can prove a simple lemma that the block adjacency matrix $\bm A_n$ is
nilpotent whenever all the subgraphs $G^{[t]}$ of $G_n$ are acyclic.

@@

\IEEEproof
Recall from the definition of the matrix $\bm A_n$ that it is block upper
triangular, reflecting causality.
Since each directed graph $G^{[t]}$ is acyclic, its corresponding adjacency
matrix $A^{[t]}$ is strictly upper triangular. As a result, $\bm A_n$ must be
upper triangular.

@@

The blocked matrix structure of the adjacency matrices presented here provide
interesting relationships between their matrix properties and the algorithmic
properties of BFS, made possible because of the reformulation of BFS as
repeated power iterations of the adjacency matrix in Algorithm~\ref{alg:bfsla}.
Note, however, that these matrices need never be instantiated for practical
computations. Rather, since Algorithm~\ref{alg:bfsla} only requires the
matrix--vector product involving the adjacency matrix, the formulation of
Algorithm~\ref{alg:bfs} provides an efficient way to exploit the block structure
of $\bm A_n$. The $\odot$ operation provides an efficient way to compute the
action of the off-diagonal products. Representing the diagonal blocks
$A^{[t]}$ as sparse matrices further reduces the cost of BFS by exploiting
latent sparsity in graphs that show up in practical applications.

@@

\IEEEproof
First, we prove that the BFS terminates in the case of acyclic evolving graphs.
Recall from Lemma~\ref{thm:nilpotent} that $\bm A_n$ is nilpotent, i.e.\
there exists some positive integer $k$ for which $\bm A_n^k = 0$.
Hence, after iteration $k$, $\bm b$ is assigned the value $\bm A_n^k \bm b = 0$.
Therefore, Algorithm~\ref{alg:bfsla} must terminate after iteration $k$.

For evolving graphs with cycles, lines 9-11 of Algorithm \ref{alg:bfsla} enforce
that the BFS visits each active node at most once. Since the $k$th block of
$\bm b$ is zeroed out if an active node has already been visited, the subgraph
traversed in the BFS cannot have cycles. Thus all that is required is the
previous result that the BFS on an acyclic graph terminates.
\endproof

@@

The complexity of the algebraic BFS of Algorithm \ref{alg:bfsla} is significantly
more complicated that that of Algorithm \ref{alg:bfs}. While the latter uses the
usual adjacency list representation for graphs, the computational cost of the
former depends critically on the actual representation of the matrices.
Furthermore, the average case analysis is complicated by the expected fill-in of
the vector $\bm b$, which influences the cost of the matrix-vector product on line 7
and the expected number of iterations of the \textbf{while} loop beginning on line6.

While a full complexity analysis is beyond the scope of this paper, it is
straightforward to present worst-case results for dense and compressed sparse
column (CSC) matrices.

\begin{lemma}[Number of iterations]
In the worst case, the number of iterations in the \textbf{while} loop of
Algorithm~\ref{alg:bfsla} is $k = O(|E|)$.
\end{lemma}

\IEEEproof
In the worst case, the BFS must traverse every active node, and only one new
active node is discovered in each iteration. The number of active
nodes is bounded above by the cardinality of the full edge set, $|E|$.
\endproof

The average case analysis for the number of iterations is considerably more
complicated and is beyond the scope of this paper.

\begin{thm}[Dense matrices]
Suppose $\bm A_n$ is represented as a dense matrix. Then the computational
complexity of Algorithm~\ref{alg:bfsla} is $O(k |V|^2)$, which in the worst case is
$O(|E| |V|^2)$.
\end{thm}

\IEEEproof
Since $\bm A_n$ is a $|V| \times |V|$ matrix, the matrix-vector product
$\bm A_n \bm b$ takes $O(|V|^2)$ operations to compute. Thus the cost of
Algorithm~\ref{alg:bfsla} is $O(k |V|^2) = O(|E| |V|^2)$ in the worst case.
\endproof

It is clear that practical implementations of the BFS should never construct
the full matrix $\bm A_n$ in memory. What happens if we use a sparse blocked
representation?

\begin{thm}[Block diagonal sparse matrices]
Suppose $\bm A_n$ is represented by a collection of compressed sparse column
matrices for each diagonal block $A^{[t]}$. Then the
computational complexity of Algorithm~\ref{alg:bfsla} is $O(k(|\tilde E| + |V|))$, which in
the worst case is $O(|E|(|\tilde E| + |V|))$.
\end{thm}

\IEEEproof
The gaxpy operation for CSC matrices costs $2n_{nz}$ flops, where $n_{nz}$ is the
number of stored values in the matrix. The cost of each diagonal subblock calculation
$A^{[t]} b_t$ is therefore $O(|E^{[t]}|)$, since $A^{[t]}$ by construction has
nonzero entries only when a static edge exists.

The off-diagonal products
$M^{[t,t\prime]} b_{t\prime}$ can be computed in $O(|V^{[t]}| + |E^{[t]}|)$ time
for all $t\prime \ge t$
since it can be implemented by the $\cdot$ operation, which constructs either a
zero vector or keeps the same vector.
The cost of checking the condition $(A^{[t]})^T b_{t^\prime} \ne 0$ is
$O(|E^{t}|)$ in the worst case since all that is required is to check whether or
not each column of $A$ is empty.
Similarly, checking the condition $A^{[t]} b_{t^\prime} \ne 0$ reduces to
checking if each row of $A$ is empty, and thus is of cost $O(|V^{t}|)$.

Thus, the cost of multiplying one block row of $\bm A^T$ (for some time $t$)
with $\bm b$ is $O(|V^{[t]}| + |E^{[t]}|)$. Summing over all times yields the
desired result.
\endproof

We can see that even implementing the BFS algebraically using CSC matrices is
insufficient to reduce the running time to linear, which can be achieved for
the adjacency list representation in Algorithm \ref{alg:bfs}.
This result strongly suggests that additional work is needed to produce true
algorithmic equivalence at the computational level.

@@

To study evolving graphs and experiment with various graph types, we have
developed EvolvingGraphs.jl~\cite{zhang15}, a software package for the creation,
manipulation, and study of evolving graphs written in Julia~\cite{bkse12}.
It is freely available online at
\begin{center}
\url{https://github.com/weijianzhang/EvolvingGraphs.jl}
\end{center}
and available with the MIT ``Expat'' license. The package contains an implementation
of the evolving graph BFS of Algorithm~\ref{alg:bfs}. \texttt{IntEvolvingGraph}, a data type in EvolvingGraphs.jl,
represents an evolving graph as adjacency lists.

We now present some simple timing data to show that our implementation of
Algorithm~\ref{alg:bfs} is indeed linear scaling in computational cost.

We generate a group of  random (directed) \texttt{IntEvolvingGraph}s
with $8,000$ active nodes and $4$ time stamps.
At each time stamp $i$, an edge is included in $G^{[i]} \in G_4$ with probability $p$.
Figure \ref{fig:time} shows the plots of number of edges against the computation time
for running Algorithm \ref{alg:bfs} in Julia. All experiments are conducted
on a Linux system with Intel(R) Core(TM) i7-3537U CPU @ $2.00$GHz.

The results show Algorithm \ref{alg:bfs} can be computed in linear time, which agrees
with Theorem \ref{thm:complexity}. A summary of the experiments is provided in Table \ref{tab:bfs}.
\begin{figure}[h]
  \centering
  \includegraphics[scale=0.44]{time.pdf}
  \caption{Experiments of running Algorithm \ref{alg:bfs}
on a group of random evolving graphs with $8,000$ active nodes and $4$ time stamps. The x-axis shows the
number of edges of each evolving graph and the y-axis shows the corresponding
computation time. The values in the x, y-axis increase linearly from left to
right and from bottom to top respectively.}
  \label{fig:time}
\end{figure}
\begin{table}[h]
\centering
\begin{tabular}{c | c | c | c }
\hline
probability $p$ & no. of edges  & time (s) &  memory (MB) \\\hline
      0.01  &         160,156      & 0.098   & 18.388  \\
      0.09  &       1,440,895        & 0.446  &  134.253  \\
      0.15   &     2,399,774         & 0.743 & 221.511 \\
      0.23   &    3,676,959          & 1.206 &   337.688 \\
      0.31   &     4,956,969          & 1.504  &  494.833 \\
\end{tabular}
\caption{the computation time and the amount of memory allocated for
computing Algorithm \ref{alg:bfs}. This is a summery of Figure \ref{fig:time}.
Each edge is included in the graph with probability $p$. }
\label{tab:bfs}
\end{table}
%Julia is a high-level, high-performance dynamic programming language for technical
%computing.
% We make particular use of
% the parameterizable type system and multiple dispatch in Julia.
% \texttt{IntMatrixList}, a data type in EvolvingGraphs.jl, represents an evolving graph
% as a sequence of sparse matrices stored in compressed sparse column (CSC) format.
% We can convert an \texttt{IntMatrixList} to its corresponding block adjacency matrix
% using the function \texttt{spmatrix}.

@@

Evolving graphs have found many applications to analyzing networks that change
over time \cite{gphe11,grihig13}. In this section, we focus specifically on
citation networks, and show that evolving graph formalism presented above can be
used to capture the dynamical structure of citation networks. Consider
the evolving graph $G_n=\langle G^{[t]} \rangle_t$ such that $G^{[t]}$ has
node set corresponding to authors active at time $t$ and directed edge set
$E^{[t]} \ni (i, j)$ representing a citation of author $j$ by author $i$ in a
publication at time $t$.

@@

We are currently investigating the use of our evolving graph BFS on citation
networks.
